---
title: "Beta-binomial prior for model dimension"
author: "Ville Karhunen"
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



Finimom employs a beta-binomial prior for model dimension $d$:

\[
\mathbb{P}(d = k | a, b) = \binom{p}{k}\frac{B(a + k, p - k + b)}{B(a, b)}, \quad a, b > 0, \quad d = 1, \dots, K,
\]

where $p$ is the number of variants and $K$ the maximum model size.
The priors of this form with $a = 1$ and $b = p^u$ with $u > 1$, are discussed in Castillo and van der Vaart (2012) and in Castillo et al. (2015).
The parameter $u$ here controls the amount of prior density for smaller models, with larger values of $u$ giving more prior mass to smaller models.

Using a linkage disequilibrium (LD) matrix from an external dataset tend to increase the false positive rate.
In our formulation, parameter $u$ provides a flexible way to adjust for this.
The default values are $u = 2$ when using in-sample LD matrix, and $u = 2.25$ when using out-of-sample LD matrix.

We demonstrate the prior for model dimension using the example dataset:

```{r setup}
library(finimom)

(p <- length(exampledata$betahat))

maxsize <- 10

a <- 1
u <- 1.5

val <- exp(sapply(seq_len(maxsize), dbb, p = p, a = a, b = p^u))
(val <- val/sum(val))

plot(val, type = "b", ylim = c(0, 1))



```


And for different values of $u$:


```{r compareu}

us <- c(1.05, 1.5, 2, 2.25)

vals <- lapply(us, function(u){
  b <- p^u
  out <- exp(sapply(1:10, dbb, a = a, p = p, b = b))
  out <- out/sum(out)
})

plot(vals[[1]], type = "b", ylim = c(0, 1))
invisible(lapply(2:4, function(i) lines(vals[[i]], type = "b", lty = i)))

```

The same on a log scale:

```{r compareulog}

plot(vals[[1]], type = "b", log = "y", ylim = range(unlist(vals)))
invisible(lapply(2:4, function(i) lines(vals[[i]], type = "b", lty = i)))


```

# References

Castillo and van der Vaart (2012). Needles and Straw in a Haystack: Posterior concentration for possibly sparse sequences. *The Annals of Statistics*.

Castillo et al. (2015). Bayesian linear regression with sparse priors. *The Annals of Statistics*.

# Session information

```{r sessioninfo}

sessionInfo()

```

