---
title: "Non-local priors"
author: "Ville Karhunen"
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}

# load required package
library(finimom)

```

Non-local densities are a family of distributions with zero density at the null parameter value.
A product inverse-moment prior for a parameter vector $\boldsymbol{\beta}$ of length $d$ has density

\[
f(\boldsymbol{\beta}|\tau, r) = \prod_{k = 1}^d\frac{\tau^{r/2}}{\Gamma(r/2)}|\beta_k|^{-(r+1)}\exp \left( -\frac{\tau}{\beta_k^2} \right)
\]


Belongs to this family.



The parameter $\tau$ controls the.
Parameter $r$ controls the tail behaviour

The choice of $\tau$ parameter is based on the suggestion by Sanyal et al. (2019).
We can obtain the corresponding quantile from the relation of the inverse-moment prior with the (inverse) Gamma distribution:

```{r invgamma}

betaq <- 0.05
prob <- 0.01

qgamma(1 - prob, shape = 0.5, scale = 1)*betaq^2

```


The correspondence of the choice $\tau = 0.0083$ is reminiscent of the Gaussian prior with $\sigma^2 = 0.15$, suggested by Wakefield (201?):

```{r comparison}

curve(dimom(x, tau = 0.0083, r = 1), -0.8, 0.8, xlab = bquote(beta), ylab = "Density")


```

Session info:

```{r sessioninfo}

sessionInfo()

```
